---
title: "How to use `MetaEgress` to generate EML document from LTER-core-metabase"
output: md_document
---
Last updated: `r format(Sys.time(), '%d %B, %Y')`

```{r echo = FALSE}
knitr::opts_chunk$set(eval = FALSE, echo = TRUE, message= FALSE, warning=FALSE)
```

This is an example workflow, taking a populated instance of LTER-core-metabase to generate complete EML document. This guide
assumes you have an installed and populated instance of
LTER-core-metabase, plus log-in credentials for an user with at least
read (or SELECT) access to this database. See the [LTER-core-metabase
Github repository](https://github.com/lter/LTER-core-metabase) for
information on installation and population of the database. If you would like to test
drive `MetaEgress`, use the database dump specifically made for this
purpose; see [this document](use_dump.md).

## Outline

- [Preparation](#preparation) 
- [Set up in R](#set-up-in-r)
- [Connect and query from LTER-core-metabase](#connect-and-query-from-lter-core-metabase)
- [Create entities](#create-entities)
- [Create EML](#create-eml)
- [Validate and write to file](#validate-and-write-to-file)


## Preparation

### Necessary components

(aside from populated LTER-core-metabase)

-   An abstract document in pandoc-compatible format. This can be a flat file, in which case the file name should be specified in `DataSet.Abstract` and `DataSet.AbstractType` set to "file", or it can be entered into LTER-core-metabase itself in plain-text or bookdown format, and the AbstractType specified accordingly.
-   A method document in pandoc-compatible format. This can be a flat file, in which case the file name should be specified in `DataSetMethod.Description` and `DataSetMethod.DescriptionType` set to "file", or it can be entered into LTER-core-metabase itself in plain-text or bookdown format, and the AbstractType specified accordingly.
-   Data files for all listed entities. File names should be exactly as listed in metabase `DataSetEntities.FileName`.

**Note on document file formats**

`MetaEgress` uses the function `set_TextType` from the EML R package to
wrap documents in XML tags. See `set_TextType` documentation to learn
more about compatible formats: `?EML::set_TextType` in R console.

### Set up a local project folder

Put data files and other documents (e.g. abstract, methods, if using files) in a folder.

Recommended approach for local files: Create a R script in this folder
and run `MetaEgress` from this script.

## Set up in R

In the project R script:

```{r}
# Load MetaEgress into R environment
library(MetaEgress)

# Set working directory to current script's directory
setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
```

## Connect and query from LTER-core-metabase

```{r}
    # View documentation
    ?get_meta

    # Use function
    metadata <-
      get_meta(
        dbname = "metabase",       # change to your DB name
        schema = "mb2eml_r",       # change to schema containing views
        dataset_ids = 1,           # change to ID or numeric vector of IDs wanted
        host = "localhost",        # change to IP address if remote host
        port = 5432,
        user = NULL,               # option to enter the username and password here to save time
        password = NULL            # if NULL, R will ask for your username and password in the console
      )
```

## Create entities

```{r}
    # View documentation
    ?create_entity_all

    # Use the function
    entities <- create_entity_all(
      meta_list = metadata,   # list returned by `get_meta`
      dataset_id = 1,         # a single dataset ID
      file_dir = NULL,        # will use current working directory if NULL
      skip_checks = T,        # if FALSE, MetaEgress will perform a congruency check comparing the names of attributes listed in your metadata and column names as appear in your data
    )
```

**Note** that even if the function runs successfully, many warnings will be generated. Use `warnings()` to see them. Most of the time the warnings will concern custom units and can safely be ignored.

## Create EML

```{r}
# View documentation
?create_EML

# Use the function
eml_document <-
  create_EML(
    meta_list = metadata,                 # list returned by `get_meta`
    entity_list = entities,               # list returned by `create_entity_all`
    dataset_id = 1                       # a singular dataset ID
)
```

## Validate and write to file

First validate EML document using the function `eml_validate` from EML R package. The input could be an EML list object or output from `create_EML`, or a XML file. 

```{r}
EML::eml_validate(EML)
```

Desired outcome is TRUE:

> [1] TRUE
> attr(,"errors")
> character(0)


Should `eml_validate` return FALSE, examine the list object returned by `create_EML`. Error messages generated by `eml_validate` could be quite cryptic and might not point to the real problem. 

Then serialize or write to XML file using the function `write_eml` from the EML R package.

```{r}
EML::write_eml(eml_document, file = "eml_document.xml")
```